---
title: "Impact of Severe Weather Events on Public Health and Economy in the United States"
author: "Isabella David"
date: "Sunday, March 08, 2015"
output: html_document
---

### Synopsis

This report analyzes the impact of different severe weather events on public health and economy in the United States from 1994 to 2011. 
Our analysis is based on data collected by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
It tracks characteristics of major storms and weather events in the U.S. from 1950 - 2011, including when and where they occur, as well as estimates of any fatalities, injuries and property damage. 
To determine the impact of storms on U.S. public health and economy, we use estimates on fatalities and injuries - on the one side - and estimates on property and crop damages - on the other side. 
We focus out attention on the period that goes from 1994 to 2011, as more recent years are most significant in terms of data availability.
Our finding is that excessive heat and tornado are most harmful with respect to population health. 
In particular, Tornado is the most hazordous climate event in terms of injuries - with more than 22,000 injuries.
Excessive heat is the most significant event in terms of fatalities - with 1,903 deaths. With respect to the impact on U.S. economy, we find that Flood, drought and hurricane/typhoon have the greatest economic consequences. 
In more details, Floods have caused the greatest property damages - more than 144 billion USD. Drought, instead, turns out to be the main cause of crop damages - with more than 13 billion USD.

### Basic settings

```{r basic, message =FALSE, warning=FALSE}
echo = TRUE  

library("R.utils")
library(dplyr)
library(ggplot2)
require(gridExtra)
```

```{r set, echo=FALSE,results="hide"}
setwd("C:/Users/ISABELLA/Documents/Coursera/Data Science Specialization/05 - Reproducible Research/WEEK3/Problem Set")
Sys.setlocale("LC_TIME", "English")
```

### Session Info
```{r }
sessionInfo()
```

### Data Processing

Read data:
```{r read, cache = TRUE}
data <- read.csv("repdata-data-StormData.csv", header = TRUE)
```

Look at data:
```{r }
str(data)
```

To reduce dataset size, we just keep columns of interest:
```{r }
storm_data <- select(data,STATE,BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

Check for missing values: 
```{r }
sum(is.na(storm_data))
```

Extract variable "year" from date format:
```{r }
storm_data <- mutate(storm_data, year = as.numeric(format(as.Date(BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y")))
```

Have a look at sample size by year:
```{r }
hist(storm_data$year, breaks = 60)
```

Select only more recent years that should be more complete:
```{r }
storm_data <- filter(storm_data, year >= 1994)
```

PROPDMGEXP and CROPDMGEXP variables need to be recoded into numerical formats according to the multiplier as indicated in the Storm Events CodeBook (H = Hundred, K = Thousand, M = Million and B = Billion). 
```{r }
levels(storm_data$PROPDMGEXP)
levels(storm_data$CROPDMGEXP)

unit <- c("", "+", "-", "?", 0:8, "h", "H", "k", "K", "m", "M", "B")
multiplier <- c(rep(0,4), 0:8, 2, 2, 3, 3, 6, 6, 9)
mult.df <- data.frame(unit, multiplier)

storm_data$PROPDMGEXP <- mult.df[match(storm_data$PROPDMGEXP, mult.df$unit),2]
storm_data$CROPDMGEXP <- mult.df[match(storm_data$CROPDMGEXP, mult.df$unit),2]
```

To get the amount of economic damages in dollars, let's multiply the number of property/crops damages by their recoded expenses ($):
```{r }
storm_data <- mutate(storm_data, PROPERTY_DAMAGE = PROPDMG * 10 ^ PROPDMGEXP, CROP_DAMAGE = CROPDMG * 10 ^ CROPDMGEXP)
```

After data processing, let's look at the dataset:
```{r }
str(storm_data)
head(storm_data)
```

### Impact on Public Health

The first part of this project asks us to find out the severest weather events in terms of population health.
Therefore, we rank the total number of fatalities by wheather event type to get the list of the top 15 severest wheather event type.

```{r }
fatalities_ranking <- 

        storm_data %>%
  	group_by(EVTYPE) %>% 
  	select(FATALITIES) %>%
  	summarise(
		FATALITIES = sum(FATALITIES)
        ) %>% 
  	arrange(desc(FATALITIES)) %>% 
  	mutate(rank = dense_rank(desc(FATALITIES))) %>%
  	filter(rank <= 15) %>%
        mutate(EVTYPE = factor(EVTYPE, levels = EVTYPE))
```

Then, we do the same for the number of injuries:

```{r }
injuries_ranking <- 

	storm_data %>%
  	group_by(EVTYPE) %>% 
  	select(INJURIES) %>%
  	summarise(
      	        INJURIES = sum(INJURIES)       
        ) %>% 
  	arrange(desc(INJURIES)) %>% 
  	mutate(rank = dense_rank(desc(INJURIES))) %>%
  	filter(rank <= 15) %>%
        mutate(EVTYPE = factor(EVTYPE, levels = EVTYPE))
```

### Impact on Economy

The second part of this project ask us to find out the severest weather events in terms of economic damages.
As in the previous section, we aggregate property/crop damages by wheather event type. Then, we ranked them to get the lists of 15 weather events that have had the severest consequences on the U.S. economy.

```{r }
property_damage_ranking <- 

        storm_data %>%
        group_by(EVTYPE) %>% 
        select(PROPERTY_DAMAGE) %>%
        summarise(
	        PROPERTY_DAMAGE = sum(PROPERTY_DAMAGE)
        ) %>% 
        arrange(desc(PROPERTY_DAMAGE)) %>% 
        mutate(rank = dense_rank(desc(PROPERTY_DAMAGE))) %>%
        filter(rank <= 15) %>%
        mutate(EVTYPE = factor(EVTYPE, levels = EVTYPE))

crop_damage_ranking <- 

        storm_data %>%
        group_by(EVTYPE) %>% 
        select(CROP_DAMAGE) %>%
        summarise(
	        CROP_DAMAGE = sum(CROP_DAMAGE)  
        ) %>% 
        arrange(desc(CROP_DAMAGE)) %>% 
        mutate(rank = dense_rank(desc(CROP_DAMAGE))) %>%
        filter(rank <= 15) %>%
        mutate(EVTYPE = factor(EVTYPE, levels = EVTYPE))
```

### Results

Let's print out the two lists with the 15 most significant storm events in terms of damages on population health: 

```{r } 
fatalities_ranking
injuries_ranking
```

and let's make a plot summarizing all these information:

```{r } 
fatalities_plot <- 
	qplot(EVTYPE, data = fatalities_ranking, weight = FATALITIES, geom = "bar") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
      geom_histogram(colour = "white", fill = "black", binwidth = 1) +
      xlab("Severe Weather Events") + 
      scale_y_continuous("Number of Fatalities") +  
      ggtitle("Number of Fatalities\n by Top 15 Severe Weather\n Events in the U.S.\n from 1994 - 2011")

injuries_plot <-
	qplot(EVTYPE, data = injuries_ranking, weight = INJURIES, geom = "bar") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
      geom_histogram(colour = "darkgreen", fill = "white", binwidth = 1) +
      xlab("Severe Weather Events") + 
      scale_y_continuous("Number of Injuries") +  
      ggtitle("Number of Injuries\n by Top 15 Severe Weather\n Events in the U.S.\n from 1994 - 2011")

grid.arrange(fatalities_plot, injuries_plot, ncol = 2)
```

From the histogram above, Tornado and Flood turn out to be the 2 severest climate events in terms of number of injuries - with 22,571 and 6,778 injuries. Excessive heat and Tornado have caused hte greatest number of fatalities - with 1,903 and 1,593 deaths from 1994 to 2011.

Finally, let's look at the 15 most significant storm events in terms of economic damages: 
 
```{r } 
property_damage_ranking
crop_damage_ranking
```

and again, let's plot the results:

```{r } 
property_damage_plot <- 
	qplot(EVTYPE, data = property_damage_ranking, weight = PROPERTY_DAMAGE/10^6, geom = "bar") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
      geom_histogram(colour = "white", fill = "darkgrey", binwidth = 1) +
      xlab("Severe Weather Events") + 
      scale_y_continuous("Property Damage [Million $]") +  
      ggtitle("Million $ Property Damage\n by Top 15 Severe Weather\n Events in the U.S.\n from 1994 - 2011")

crop_damage_plot <-
	qplot(EVTYPE, data = crop_damage_ranking, weight = CROP_DAMAGE/10^6, geom = "bar") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
      geom_histogram(colour = "white", fill = "brown", binwidth = 1) +
      xlab("Severe Weather Events") + 
      scale_y_continuous("Crop Damage [Million $]") +  
      ggtitle("Million $ Crop Damage\n by Top 15 Severe Weather\n Events in the U.S.\n from 1994 - 2011")

grid.arrange(property_damage_plot, crop_damage_plot, ncol = 2)
```

In terms of property damages, we show that Floods and Hurricane/Typhoon have been the most severe weather events - with more than 144 and 69 billion USD, respectively. 
We also show that Drought and Flood represent the top 2 causes of crop damages - with more than 13 and 5 billion USD, respectively.

### Conclusion
Our finding is that across the United States from 1994 to 2011, Excessive heat and Tornado had the greatest impact on most population health - while Flood, Hurricane/typhoon and Drought had the greatest economic consequences.




